Locally-Oriented Programming:\\ A Simple Programming Model for Stencil-Based Computations on Multi-Level Distributed Memory Architectures

### 2015-02-09 papers/LOPe/transformations.tex
Finished beta-level changes.

This section briefly describes how LOPe extensions to Fortran have been implemented as
source-to-source transformations via rewrite rules and strategies in StrategoXT.

Transformations are accomplished in three phases:
1. Parsing to produce an AST nodes representated as ATerms;
2. Transformations on the AST nodes represented as rewrite rules and strategies in Stratego/XT; and
3. Pretty printing of transformed nodes using the pretty printing facilites of BoxLib.

@@

A key component of the LOPe transformations is the targeting of a LOPe \texttt{CONCURRENT}
procedure for a particular hardware (and software) architecture.  The execution target can be one
of several choices, including serial execution by the current process (and hardware) via inlining
of the function, parallel execution by inlining with OpenMP compiler directives, or parallel
execution by a heterogeneous processing elements with a language like OpenCL, for example.

For this work we have developed rewrite strategies and rules in Stratego/XT to rewrite Fortran AST
nodes to C AST nodes (extended with necessary OpenCL keywords).  The C AST ATerms have mostly a
one-to-one correspondence with Fortran terms.  A \texttt{CONCURRENT} procedure is transformed to an
OpenCL kernel, Fortran formal parameters are transformed to C parameters (with a direct mapping of
types), and local Fortran variable declarations are rewritten as C declarations.  Similarly,
Fortran executable statements are rewritten as C executable AST nodes.  Mostly the only
complication is mapping the LOPe local array index view to the global C index space.  This
translation is facilitated by a Fortran symbol table that stores array shape and halo size
information.

Once all rewriting passes have been made to transform from LOPe Fortran ATerms to C ATerms, the
C ATerms are output as an OpenCL file using a C pretty printer.

\subsection{Transformations at the Call Site}

Transformations at a LOPe procedure call site are more difficult, though technically straight forward.
The Fortran function call must be transformed to a call to run the OpenCL kernel (produced as described
above).  This is facilitated by the use of the ForOpenCL library that provides Fortran bindings to the
OpenCL runtime\cite{foropencl}.  But the usage requires the declaration of extra variables,
allocation of memory on the OpenCL device (subimage), transfer of memory, marshalling of kernel
arguments, and synchronization.

These transformations are accomplished using several rewrite stages using Stratego/XT strategies:
1. A symbol table is produced in the first pass to store information related to arrays including,
array shape, halo size, and allocation status;
2. Additional variables are declared that are used to maintain the OpenCL runtime state, including
the OpenCL device (as obtained from a \texttt{GET\_SUBIMAGE} call), the OpenCL kernel (produced as
described above), and OpenCL variables used to marshall kernel arguments; and
3. All CAFe code related to subimage usage is lowered to standard Fortran with calls to the
ForOpenCL library as needed.  The latter step includes the allocation of OpenCL device memory,
transfer of memory to and from the OpenCL device and the executing Fortran process (image) as
indicated in the program source by the single bracket \texttt{[subimage]} notation, marshalling
of kernel argument, running of the OpenCL kernel, and synchronization.

Though not yet available, similar rewrite strategies are planned for targeting programming models
other than OpenCL including parallel execution with OpenMP directives.  In addition, simple serial
execution with function inlining (if desired) will be performed by the regular Fortran compiler once
all CAFe and LOPe code has been lowered to standard Fortran.

>>>>

This section briefly describes how LOPe extensions to Fortran have been implemented as
source-to-source transformations via rewrite rules (for expressing basic transformations) and
rewriting strategies (for controlling the application of the rewrite rules).  A LOPe file is
transformed to Fortran and OpenCL files through generative programming techniques using
Stratego/XT tools\cite{Bravenboer200852} and is accomplished in three phases: (1) parsing to
produce a LOPe Abstract Syntax Tree (AST) represented in the Annotated Term
Format (ATerm\cite{DBLP:journals/spe/BrandJKO00}); (2) transformations of LOPe AST nodes to Fortran
and C AST nodes; and (3) pretty-printing to the base Fortran and OpenCL languages.

The foundation of LOPe is the syntax definition of the base language expressed in
SDF (Syntax Definition Formalism) as part of the Open Fortran Project (OFP)\cite{OFP:git:url}.
LOPe is defined in a separate SDF module that extends the Fortran 2008 language standard with 15
context-free syntax rules.  Parsing is implemented in Stratego/XT by a scannerless generalized-LR
parser and the conversion of transformed AST nodes to text is accomplished with simple
pretty-printing rules.

@@

A key component of code generation for LOPe is the targeting of a LOPe \texttt{CONCURRENT}
procedure for a particular hardware architecture.  The execution target can be one
of several choices, including serial execution by the current process via inlining
of the function, parallel execution by inlining with OpenMP compiler directives, or parallel
execution by heterogeneous processing elements with a language like OpenCL.

For this work we have developed rewrite rules and strategies in Stratego/XT to rewrite Fortran AST
nodes to C AST nodes (extended with necessary OpenCL keywords).  The C AST ATerms have mostly a
one-to-one correspondence with Fortran terms: a \texttt{CONCURRENT} procedure is transformed to an
OpenCL kernel; Fortran formal parameters are transformed to C parameters (with a direct mapping of
types); and local Fortran variable declarations are rewritten as C declarations.  Similarly,
Fortran executable statements are rewritten as C statements.  The only
minor complication is mapping the LOPe local, array index view to the global C index space.  This
translation is facilitated by a Fortran symbol table that stores array shape and halo size
information.

\subsection{Transformations of the Calling Site}

Transformations of a LOPe procedure call site are more difficult, though technically straight
forward.  The Fortran function call must be transformed to a call to run the OpenCL kernel
(generated as described above).  This is facilitated by use of the ForOpenCL library which
provides Fortran bindings to the OpenCL runtime\cite{Sottile:2013:FTE:2441516.2441520}.  However,
this usage requires the declaration of extra variables, allocation of memory on the OpenCL device
(subimage), transfer of memory, marshalling of kernel arguments, and synchronization.

These transformations are accomplished using several rewrite stages using Stratego/XT strategies:
(1) a symbol table is produced in the first pass to store information related to arrays including,
array shape, halo size, and allocation status;
(2) additional variables are declared that are used to maintain the OpenCL runtime state, including
the OpenCL device, the OpenCL kernel, and OpenCL variables used to marshall kernel arguments; and
(3) all CAFe code related to subimage usage is desugared to standard Fortran with calls to the
ForOpenCL library as needed.
%%The latter step includes the allocation of OpenCL device memory, transfer of memory to and from the OpenCL device, marshalling of kernel arguments, running of the OpenCL kernel, and synchronization.

Though not yet available, similar rewrite strategies are planned for targeting programming models
other than OpenCL including parallel execution with OpenMP directives.  In addition, simple serial
execution with function inlining (if desired) will be performed by the regular Fortran compiler once
all CAFe and LOPe code has been desugared (lowered) to standard Fortran.
